General limits

### 2017-09-08 papers/talk-7/paper-7.tex
reworked colimit section

\maketitle
\section*{Introduction}
We have seen three special structures in categories,
the product, the pullback, and the equalizer.
As an example, the binary product is constructed in the following way.
We start with two objects. Then an object $U$ with maps to these objects is the product,
if it exists an satisfies the universal property for products, that is, all other such objects $V$
factor through $U$ uniquely.\\[1em]
\begin{minipage}{.35\linewidth}
  \begin{tikzcd}
    \bullet &&\bullet
  \end{tikzcd}
\end{minipage}%
\begin{minipage}{.35\linewidth}
  \begin{tikzcd}
    & U \arrow{dl} \arrow{dr} &\\
    \bullet &&\bullet
  \end{tikzcd}
\end{minipage}%
\begin{minipage}{.35\linewidth}
  \begin{tikzcd}
    & V \arrow[bend right=40]{ddl} \arrow[bend left=40]{ddr} \arrow[dashed]{d} &\\
    & U \arrow{dl} \arrow{dr} &\\
    \bullet &&\bullet
  \end{tikzcd}
\end{minipage}
\\[1em]
The pullback and the equalizer are constructed similarly.
They all have an input diagram, and we call an object $U$ with maps into this diagram
product/pullback/equalizer if it satisfies the universal property.
\\[1em]
\begin{minipage}{.35\linewidth}
  \begin{tikzcd}
    & V \arrow[bend right=40]{ddl} \arrow[bend left=40]{ddr} \arrow[dashed]{d} &\\
    & U \arrow{dl} \arrow{dr} &\\
    \bullet &&\bullet
  \end{tikzcd}
\end{minipage}%
\begin{minipage}{.35\linewidth}
  \begin{tikzcd}
    V
    \arrow[bend left=40]{drr}
    \arrow[bend right=40]{ddr}
    \arrow[dashed,swap]{dr}
    && \\
    &U
    \arrow{r}
    \arrow{d}
    &\bullet \arrow{d} \\
    & \bullet \arrow{r} & \bullet
  \end{tikzcd}
\end{minipage}%
\begin{minipage}{.3\linewidth}
  \begin{tikzcd}
    V \arrow[dashed]{dd} \arrow{dr}&& \\
    & \bullet \arrow[shift left]{r} \arrow[shift right]{r} & \bullet \\\
    U \arrow{ur} &&
  \end{tikzcd}
\end{minipage}\\[2em]
%% TODO: possibly give further applications of limits too
%% TODO: maybe some more intuition what limits are about

\section{Definition}
Products, pullbacks, and equalizers are instances of a more general concept called
limit which expresses the common pattern between the examples above.
The goal is to make this abstract common pattern more formal
and establish relations between concrete instances of it.
We start with the definition of a limit.\\

\begin{definition}
  A \textbf{limit} is a universal cone over a diagram.
\end{definition}

In the following section we explain this definition. We need to clarify three things
in particular.
First we need to define the concept of a diagram more formally.
Then we define what a cone over a diagram is. Lastly we state what it means for a cone
to be universal.
We start with diagrams. They are the initial entities of which we take the limit.
In the examples above the diagrams
are the various collections of objects (and arrows) that we start with
and over which the object $U$ has maps.

\begin{definition}
  Let $\bm{I}, \mathscr{C}$ be categories.
  A \textbf{diagram} is a functor $D \from \bm{I} \to \mathscr{C}$.
  The category $\bm{I}$ is called the \textbf{shape} of the diagram.
\end{definition}
\begin{example}
  \label{diagramexample}
  This is an example of a diagram we use to build a pullback. \\[1cm]
  \begin{minipage}{0.3\linewidth}
    \hspace{2cm}
    \begin{tikzcd}
      & \bullet \arrow{d} \\
      \bullet\arrow{r} & \bullet
    \end{tikzcd}\\
  \end{minipage}%
  \begin{minipage}{0.2\linewidth}
    $\qquad\quad\overset{D}{\Longrightarrow}$
  \end{minipage}%
  \begin{minipage}{0.1\linewidth}
    \begin{tikzpicture}
      \draw(1,0) circle (1.5cm);
      \begin{tikzcd}
        & \bullet \arrow{d} \\
        \bullet\arrow{r} & \bullet
      \end{tikzcd}
    \end{tikzpicture}
  \end{minipage}\\[2em]
  \begin{minipage}{0.3\linewidth}
    \hspace{2cm}
    $\qquad \bm I$
  \end{minipage}%
  \begin{minipage}{0.2\linewidth}
    \
  \end{minipage}%
  \begin{minipage}{0.3\linewidth}
    \hspace{2cm}
    $\qquad \mathscr C$
  \end{minipage}\\[1em]
  We did not give names to the objects as we are not concerned with them.
\end{example}

@@

%%TODO: fix this section
\section{Colimits}
To define colimits we dualize the concept of limits.
\begin{definition}
  Let $D : \bm I \to \mathscr C$ be a diagram.
  A \textbf{cocone} $U$ over $D$ is an object $U \in \mathscr{C}$
  with maps $D_I \overset{ f_I}{\rightarrow} U,\ \forall\ I \in \bm{I}$, such that
  $\forall\ I \overset{m}{\to} J \in \bm{I}$, the following triangle commutes:
  \[
    \begin{tikzcd}
      D_I \arrow[rr, "D(m)"]\arrow[dr, swap, "f_I"] & & D_J \arrow{dl}{f_J}\\
      & U  &
    \end{tikzcd}
  \]
  That is $f_I = f_J \circ D(m)$.
\end{definition}
A cone has maps from the vertex into the diagram,
while a cocone has maps from the diagram into the vertex.
\begin{definition}
  A cocone $(D(I)\overset{f_I}{\to} U)_{I\in \bm{I}}$ is \textbf{universal}, if
  for all other cones $(D(I) \overset{g_I}{\to} V)_{I\in \bm{I}}$ there is a unique
  map $h \from U \to V$, such that for all $I \in \bm{I}$
  the following triangles commute.
  \[
    \begin{tikzcd}
      D_I \arrow[dr, swap, "g_I"] \arrow{r}{f_I} & U \arrow{d}{h} \\
      & V
    \end{tikzcd}
  \]
  That is $g_i = f_i \circ h$.
\end{definition}

\begin{definition}
  A \textbf{colimit} is a universal co-cone over a diagram.
\end{definition}
%% TODO: complete definition

\[
    \begin{tikzcd}
      & D_K \arrow{d}{D_n}
      \arrow[bend left=30,color=blue]{ddr}{f_K}
      \arrow[bend left=50,color=red]{dddrr}{g_K} & & \\
      D_I \arrow{r}{D_m}
      \arrow[bend right=30,color=blue]{drr}{f_I}
      \arrow[bend right=40,color=red]{ddrrr}{g_I} &
      D_J \arrow[bend left=20,swap, color=blue]{dr}{f_J}
      \arrow[bend left=40,color=red]{ddrr}{g_J} & & \\
      & & {\color{blue}U} \arrow[dashed]{dr}{h}& \\
      & & & {\color{red}V}
    \end{tikzcd}
\]
%% TODO: example drawing

>>>>

\maketitle
\section*{Introduction}
We have seen three special structures in categories,
the product, the pullback, and the equalizer.
As an example, the binary product is constructed in the following way.
We start with two objects. Then an object $U$ with maps to these objects is the product,
if it exists an satisfies the universal property for products, that is, all other such objects $V$
factor through $U$ uniquely.\\[1em]
\begin{minipage}{.26\linewidth}
  \begin{tikzcd}
    \bullet &&\bullet
  \end{tikzcd}
\end{minipage}%
\begin{minipage}{.1\linewidth}
  $\rightsquigarrow \quad$
\end{minipage}%
\begin{minipage}{.3\linewidth}
  \begin{tikzcd}
    & U \arrow{dl} \arrow{dr} &\\
    \bullet &&\bullet
  \end{tikzcd}
\end{minipage}%
\begin{minipage}{.1\linewidth}
  $\rightsquigarrow \quad$
\end{minipage}%
\begin{minipage}{.35\linewidth}
  \begin{tikzcd}
    & V \arrow[bend right=40]{ddl} \arrow[bend left=40]{ddr} \arrow[dashed]{d} &\\
    & U \arrow{dl} \arrow{dr} &\\
    \bullet &&\bullet
  \end{tikzcd}
\end{minipage}
\\[1em]
The pullback and the equalizer are constructed similarly.
They all have an input diagram, and we call an object $U$ with maps into this diagram
product/pullback/equalizer if it satisfies the universal property.
\\[1em]
\begin{minipage}{.35\linewidth}
  \begin{tikzcd}
    & V \arrow[bend right=40]{ddl} \arrow[bend left=40]{ddr} \arrow[dashed]{d} &\\
    & U \arrow{dl} \arrow{dr} &\\
    \bullet &&\bullet
  \end{tikzcd}
\end{minipage}%
\begin{minipage}{.35\linewidth}
  \begin{tikzcd}
    V
    \arrow[bend left=40]{drr}
    \arrow[bend right=40]{ddr}
    \arrow[dashed,swap]{dr}
    && \\
    &U
    \arrow{r}
    \arrow{d}
    &\bullet \arrow{d} \\
    & \bullet \arrow{r} & \bullet
  \end{tikzcd}
\end{minipage}%
\begin{minipage}{.3\linewidth}
  \begin{tikzcd}
    V \arrow[dashed]{dd} \arrow{dr}&& \\
    & \bullet \arrow[shift left]{r} \arrow[shift right]{r} & \bullet \\\
    U \arrow{ur} &&
  \end{tikzcd}
\end{minipage}\\[2em]
%% TODO: possibly give further applications of limits too
%% TODO: maybe some more intuition what limits are about

\section{Definition}
Products, pullbacks, and equalizers are instances of a more general concept called
limit which expresses the common pattern between the examples above.
The goal is to make this abstract common pattern more formal
and establish relations between concrete instances of it.
We start with the definition of a limit.
\begin{definition}
  A \textbf{limit} is a universal cone over a diagram.
\end{definition}

In the following section we explain this definition. We need to clarify three things
in particular.
First we need to define the concept of a diagram more formally.
Then we define cones over a diagram. Lastly we state what it means for a cone
to be universal.
We start with diagrams. They are the initial entities of which we take the limit.
In the examples above the diagrams
are the various collections of objects (and arrows) that we start with
and over which the object $U$ has maps.

\begin{definition}
  Let $\bm{I}, \mathscr{C}$ be categories.
  A \textbf{diagram} is a functor $D \from \bm{I} \to \mathscr{C}$.
  The category $\bm{I}$ is called the \textbf{shape} of the diagram.
\end{definition}
\begin{example}
  \label{diagramexample}
  This is an example of the diagram used for the pullback. \\[1cm]
  \begin{minipage}{0.3\linewidth}
    \hspace{2cm}
    \begin{tikzcd}
      & \bullet \arrow{d} \\
      \bullet\arrow{r} & \bullet
    \end{tikzcd}\\
  \end{minipage}%
  \begin{minipage}{0.2\linewidth}
    $\qquad\quad\overset{D}{\Longrightarrow}$
  \end{minipage}%
  \begin{minipage}{0.1\linewidth}
    \begin{tikzpicture}
      \draw(1,0) circle (1.5cm);
      \begin{tikzcd}
        & \bullet \arrow{d} \\
        \bullet\arrow{r} & \bullet
      \end{tikzcd}
    \end{tikzpicture}
  \end{minipage}\\[2em]
  \begin{minipage}{0.3\linewidth}
    \hspace{2cm}
    $\qquad \bm I$
  \end{minipage}%
  \begin{minipage}{0.2\linewidth}
    \
  \end{minipage}%
  \begin{minipage}{0.3\linewidth}
    \hspace{2cm}
    $\qquad \mathscr C$
  \end{minipage}\\[1em]
  We did not give names to the objects as we are not concerned with them.
\end{example}

@@

%%TODO: fix this section
\section{Colimits}
 Colimits are the dual of limits.
The motivation to dualize limits is to obtain dualized versions of theorems for limits.
Many concepts in mathematics turn out to be colimits and it is
therefore useful to have a general definition for them as well.
\begin{definition}
  A \textbf{colimit} is a universal cocone over a diagram.
\end{definition}
A cocone is the dualized version of a cone.
Recall that a cone has maps from the vertex into the diagram. Dually,
a cocone has maps from the diagram into the vertex.
\begin{definition}
  Let $D : \bm I \to \mathscr C$ be a diagram.
  A \textbf{cocone} $U$ over $D$ is an object $U \in \mathscr{C}$
  with maps $D_I \overset{ f_I}{\rightarrow} U,\ \forall\ I \in \bm{I}$, such that
  $\forall\ I \overset{m}{\to} J \in \bm{I}$, the following triangle commutes:
  \[
    \begin{tikzcd}
      D_I \arrow[rr, "D(m)"]\arrow[dr, swap, "f_I"] & & D_J \arrow{dl}{f_J}\\
      & U  &
    \end{tikzcd}
  \]
  That is $f_I = f_J \circ D(m)$.
\end{definition}
The universality condition is also dualized as expected.
For limits the condition is the existence of a unique map from
any other cone to the limit cone.
For colimits the condition is therefore dually the existence of a unique map
from the colimit to any other cocone.
\begin{definition}
  A cocone $(D(I)\overset{f_I}{\to} U)_{I\in \bm{I}}$ is \textbf{universal}, if
  for all other cocones $(D(I) \overset{g_I}{\to} V)_{I\in \bm{I}}$ there is a unique
  map $h \from U \to V$, such that for all $I \in \bm{I}$
  the following triangles commute.
  \[
    \begin{tikzcd}
      D_I \arrow[dr, swap, "g_I"] \arrow{r}{f_I} & U \arrow{d}{h} \\
      & V
    \end{tikzcd}
  \]
  That is $g_i = f_i \circ h$.
\end{definition}
\begin{example}
\[
    \begin{tikzcd}
      & D_K \arrow{d}{D_n}
      \arrow[bend left=30,color=blue]{ddr}{f_K}
      \arrow[bend left=50,color=red]{dddrr}{g_K} & & \\
      D_I \arrow{r}{D_m}
      \arrow[bend right=30,color=blue]{drr}{f_I}
      \arrow[bend right=40,color=red]{ddrrr}{g_I} &
      D_J \arrow[bend left=20,swap, color=blue]{dr}{f_J}
      \arrow[bend left=40,color=red]{ddrr}{g_J} & & \\
      & & {\color{blue}U} \arrow[dashed]{dr}{h}& \\
      & & & {\color{red}V}
    \end{tikzcd}
  \]
\end{example}

\begin{example}
  We now verify that our three examples are indeed instances of the general colimit definition.
  The diagram for the colimit is the dualized diagram of the limit.
  For coproducts the diagram is therefore the same as for products.
  We just construct a cocone instead of a cone.
  \[
    \begin{tikzcd}
      \bullet \arrow{dr} \arrow[bend right=30]{ddr}&& \bullet \arrow{dl} \arrow[bend left= 30]{ddl}\\
      & U \arrow[dashed]{d} & \\
      & V &
    \end{tikzcd}
  \]
  It is easy to see that the commuting condition for universal cocones is exactly the commuting condition
  for coproducts.\\
  Next we study the coequalizer.
  We start with the diagram
  \[
    \begin{tikzcd}
      X \arrow[shift left]{r}{f} \arrow[shift right, swap]{r}{g} & Y
    \end{tikzcd}
  \]
  We construct a cocone over it
  \[
    \begin{tikzcd}
      X \arrow[shift left]{r}{f} \arrow[shift right, swap]{r}{g} \arrow[swap]{d}{p_X} & Y \arrow[bend left=30]{dl}{p_Y}\\
      U
    \end{tikzcd}
  \]
  Because of the commuting condition for cocones we have
  $p_1 = p_2 \circ f = p_2 \circ g$.
  Since $p_1$ is uniquely determined by composition we leave the arrow implicit and do not draw it for all further cones.
  If $U$ is a colimit, then we have
  \[
    \begin{tikzcd}
      X \arrow[shift left]{r}{f} \arrow[shift right, swap]{r}{g} & Y \arrow{r}{p_Y} \arrow{dr}{q_Y}& U \arrow[dashed]{d}{h} \\
      && V
    \end{tikzcd}
  \]
  The commuting condition says that the diagram commutes.
  Specifically we have $h \circ p \circ f= q \circ f$ and $h \circ p \circ g = q \circ g$ which is exactly
  the commuting condition for coequalizers.\\
  We have similarly that a pushout is a colimit of the diagram \begin{tikzcd} I & J \arrow{l}{f} \arrow{r}{g} & K \end{tikzcd}
  by leaving out the projection map $p_J$.
\end{example}
